An exposition on the propriety of restricted Boltzmann machines

### 2016-08-22 writing/draft.Rmd
update intro and refs

source("helpers/functions.R")
```
#Introduction

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM). (TODO: deep learning, deep RBM references.) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer. These models are typically used with success in classification of images. (TODO: classification references.) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse. In this paper, we attept to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology. (TODO: statistical model definition) Through our study, we find the RBM model to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data is generated. Specifically, these models often fail to generate data that resembles realistic data. Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space. In practice, this is seen when a single pixel change in an image results in a wildly different classification (TODO: add ref). We investigate these phenomena in section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, a computationally intractible situation is created quickly when fitting with either maximum likelihood or through Bayesian methods. However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs. In section \ref{bayesian-model-fitting} we try three Bayesian modelling techniques, each with the goal of avoiding parts of the parameter space that yield improper models. As could be expected, with computational complexity comes an increase in accuracy, but at the cost of feasibility. Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution. Thus, there is no smoothing achieved with a RBM that is fit using a rigourous method. We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there are two accompanying appendices with auxillary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.

#Background

##Restricted Boltzmann machines

@@

###Joint distribution
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then we posit a parametric form for probabilities corresponding to the state of each node taking the value of $1$.

@@

Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be infeasible to calculate, making it impossible to estimate the model exactly. An approximation can be achieved using Gibbs sampling, which is a natural fit due to the restricted architecture of the RBM. Specifically, the conditional independence of each layer given the other allows for simulation of the main and hidden layers.

###Connection to Deep Learning
RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, the method's proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired architecture is created.

##Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, discuss how to detect them in the RBM, and point out their relationship to one another.

###Near-degeneracy
In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}. In other words, if random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$ then the model is near-degenerate, as defined by @handcock2003assessing.

@@

###Instability
@schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses. This can be characterized of as excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. @schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.

@@

###Uninterpretability
@kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model. This measure concerns the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$. Using this concept of dependence in a model, it is possible to understand what conditions lead to uninterpretability in a model versus those that guarantee interpretable models. In this method, a model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

@@

#Explorations

@@

##Tiny example
To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

###Impropriety three ways

@@

###Data coding to mitigate degeneracy
Multiple encodings of the binary variables are possible. For example, we could allow $\mathcal{H} \subset \{0,1\}^H, \mathcal{V} \subset \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \subset \{-1,1\}^H, \mathcal{V} \subset \{-1,1\}^V$. This will result in $t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes. To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restructed convex hull of $\{t(\boldsymbol x)\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.

For the small two node example, the binary encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume compared to unrestricted space and the negative encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume compared to unrestricted space. This notion of lost volume can be helpful to conceptualize how difficult it will be for the mean parameterized vector $\mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near degeneracy. In fact, if we look at the ratio of volume within a convex hull versus the volume of an unrestricted hull, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increses, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy. Additionally, it appears that the $\{-1,1\}$ encoding suffers slightly less from this problem.

@@

##Exploring manageable examples
```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
#reshape data functions
plot_data <- function(res, grid = FALSE) {

@@

#Bayesian model fitting
To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy, instability, and uninterpretability. Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.

@@

The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step. All three priors do require choosing the values of hyperparameters. In each case, we have chosed the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, making it less appealing. The values used for the hyperparameters are presented in table \ref{tab:hyperparam}.

@@

Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section. The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution. From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data. However, this model requires a marginalization step, which is infeasible for a model of any real size.

@@

#Discussion

>>>>

source("helpers/functions.R")
```

# Introduction

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer. These models are typically used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse. In this paper, we attempt to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology. Through our study, we find the RBM model to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data is generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. In the case of RBMs, these models often fail to generate data that resembles realistic data. Thus, providing an unsatisfactory conceptualization of the data generation process. Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space. In practice, this is seen when a single pixel change in an image results in a wildly different classification. These model impropriety measures has been documented in other deep architectures recently as well [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, a computationally intractable situation is created quickly when fitting with either maximum likelihood or through Bayesian methods. However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs. In section \ref{bayesian-model-fitting} we try three Bayesian modelling techniques, each with the goal of avoiding parts of the parameter space that yield improper models. As could be expected, with computational complexity comes an increase in accuracy, but at the cost of feasibility. Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution. Thus, there is no smoothing achieved with a RBM that is fit using a rigorous method. We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there are two accompanying appendices with auxiliary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.

# Background

## Restricted Boltzmann machines

@@

### Joint distribution
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then we posit a parametric form for probabilities corresponding to the state of each node taking the value of $1$.

@@

Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be unfeasible to calculate, making it impossible to estimate the model exactly. An approximation can be achieved using Gibbs sampling, which is a natural fit due to the restricted architecture of the RBM. Specifically, the conditional independence of each layer given the other allows for simulation of the main and hidden layers.

### Connection to Deep Learning
RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, the method's proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired architecture is created.

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, discuss how to detect them in the RBM, and point out their relationship to one another.

### Near-degeneracy
In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}. In other words, if random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$ then the model is near-degenerate, as defined by @handcock2003assessing.

@@

### Instability
@schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses. This can be characterized of as excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. @schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.

@@

### Uninterpretability
@kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model. This measure concerns the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$. Using this concept of dependence in a model, it is possible to understand what conditions lead to uninterpretability in a model versus those that guarantee interpretable models. In this method, a model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

@@

# Explorations

@@

## Tiny example
To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

### Impropriety three ways

@@

### Data coding to mitigate degeneracy
Multiple encodings of the binary variables are possible. For example, we could allow $\mathcal{H} \subset \{0,1\}^H, \mathcal{V} \subset \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \subset \{-1,1\}^H, \mathcal{V} \subset \{-1,1\}^V$. This will result in $t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes. To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restricted convex hull of $\{t(\boldsymbol x)\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.

For the small two node example, the binary encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume compared to unrestricted space and the negative encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume compared to unrestricted space. This notion of lost volume can be helpful to conceptualize how difficult it will be for the mean parameterized vector $\mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near degeneracy. In fact, if we look at the ratio of volume within a convex hull versus the volume of an unrestricted hull, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy. Additionally, it appears that the $\{-1,1\}$ encoding suffers slightly less from this problem.

@@

## Exploring manageable examples
```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
#reshape data functions
plot_data <- function(res, grid = FALSE) {

@@

# Bayesian model fitting
To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy, instability, and uninterpretability. Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.

@@

The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step. All three priors do require choosing the values of hyperparameters. In each case, we have chosen the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, making it less appealing. The values used for the hyperparameters are presented in table \ref{tab:hyperparam}.

@@

Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section. The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution. From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data. However, this model requires a marginalization step, which is unfeasible for a model of any real size.

@@

# Discussion
